---
title: "Making R Suck Less:\n\nAn Ode to `...`"
author: Fred Callaway
date: May 3, 2024
output: slidy_presentation
---

```{r, include=FALSE}
VERSIONS = c('v5.6')
knitr::opts_chunk$set(
    fig.path="figs/",
    warning=FALSE, message=FALSE, fig.width=3, fig.height=3, fig.align="center"
)
```

```{r echo=FALSE}
library(tidyverse)
library(ggbeeswarm)
library(patchwork)

# general purpose plot aesthetics

kable = knitr::kable

default_palettes = list(
  c("dodgerblue"),
  c("dodgerblue", "coral"),
  RColorBrewer::brewer.pal(6, "Set1")
)

options(
  "max.print" = 100,
  "width" = 150,
  "ggplot2.discrete.colour" = default_palettes,
  "ggplot2.discrete.fill" = default_palettes
)

theme_set(theme_bw(base_size = 14))
theme_update(
  panel.grid.major.x = element_blank(),
  panel.grid.minor.x = element_blank(),
  panel.grid.minor.y = element_blank(),
  # panel.grid.major.y = element_blank(),
  panel.grid.major.y = element_line(color="#EDEDED"),
  strip.background = element_blank(),
  strip.text.x = element_text(size=12),
  strip.text.y = element_text(size=12),
  legend.position="right",
  panel.spacing = unit(1, "lines"),
)

update_geom_defaults("vline", list(linetype="dashed"))
update_geom_defaults("abline", list(linetype="dashed"))
update_geom_defaults("hline", list(linetype="dashed"))
update_geom_defaults("line", list(linewidth = 1))
update_geom_defaults("smooth", list(color="black"))
update_geom_defaults("pointrange", list(size=.3))

# our example dataset
data("iris")
```

# The typical way

```{r}
iris %>% ggplot(aes(Species, Sepal.Width, color=Species)) +
  stat_summary(fun.data=mean_cl_boot, geom="pointrange") +
  geom_quasirandom(alpha=0.3) +
  scale_colour_manual(values=c(
    setosa = "#842EBD",
    versicolor = "#CA26E2",
    virginica = "#EB75F5"
  )) +
  theme(legend.position="none")

```

# The typical way


```{r}

iris %>% ggplot(aes(Species, Sepal.Length, color=Species)) +
  stat_summary(fun.data=mean_cl_boot, geom="pointrange") +
  geom_quasirandom(alpha=0.3) +
  scale_colour_manual(values=c(
    setosa = "#842EBD",
    versicolor = "#CA26E2",
    virginica = "#EB75F5"
  )) +
  theme(legend.position="none")
```

# The typical way

```{r}
iris %>% ggplot(aes(Species, Petal.Width, color=Species)) +
  stat_summary(fun.data=mean_cl_boot, geom="pointrange") +
  geom_quasirandom(alpha=0.3) +
  scale_colour_manual(values=c(
    setosa = "#842EBD",
    versicolor = "#CA26E2",
    virginica = "#EB75F5"
  )) +
  theme(legend.position="none")
```

# The typical way sucks

```{r eval=FALSE}
iris %>% ggplot(aes(Species, Sepal.Width, color=Species)) +
  stat_summary(fun.data=mean_cl_boot, geom="pointrange") +
  geom_quasirandom(alpha=0.3) +
  scale_colour_manual(values=c(
    setosa = "#842EBD",
    versicolor = "#CA26E2",
    virginica = "#EB75F5"
  )) +
  theme(legend.position="none")

iris %>% ggplot(aes(Species, Sepal.Length, color=Species)) +
  stat_summary(fun.data=mean_cl_boot, geom="pointrange") +
  geom_quasirandom(alpha=0.3) +
  scale_colour_manual(values=c(
    setosa = "#842EBD",
    versicolor = "#CA26E2",
    virginica = "#EB75F5"
  )) +
  theme(legend.position="none")

iris %>% ggplot(aes(Species, Petal.Width, color=Species)) +
  stat_summary(fun.data=mean_cl_boot, geom="pointrange") +
  geom_quasirandom(alpha=0.3) +
  scale_colour_manual(values=c(
    setosa = "#842EBD",
    versicolor = "#CA26E2",
    virginica = "#EB75F5"
  )) +
  theme(legend.position="none")
```

# Using a function with `{{var}}`

```{r, fig.width=9}
plot_species = function(data, yvar) {
    data %>% ggplot(aes(Species, {{yvar}}, color=Species)) +
        stat_summary(fun.data=mean_cl_boot, geom="pointrange") +
        geom_quasirandom(alpha=0.3) +
        scale_colour_manual(values=c(
            setosa = "#842EBD",
            versicolor = "#CA26E2",
            virginica = "#EB75F5"
    )) +
        theme(legend.position="none")
}

iris %>% plot_species(Petal.Width) +  # library(patchwork)
iris %>% plot_species(Petal.Length) +
iris %>% plot_species(Petal.Width)
```

# Using a function with `{{var}}`

[blog post](https://www.tidyverse.org/blog/2019/06/rlang-0-4-0/#a-simpler-interpolation-pattern-with-)

```{r}
mean_by = function(data, var, by) {
  data %>%
    group_by({{ by }}) %>%
    summarise("{{var}}" := max({{ var }}, na.rm = TRUE))
}

iris %>% mean_by(Petal.Width, by=Species) %>% kable
```

# Arbitrary arguments with `...`

```{r fig.width=9}
plot_relationship = function(data, ...) {
    data %>% ggplot(aes(...)) +
        geom_point(alpha=0.5) +
        stat_smooth(method="lm",  alpha=0.2)
}

iris %>% plot_relationship(Petal.Width, Petal.Length) +
iris %>% plot_relationship(Petal.Width, Sepal.Width) +
iris %>% plot_relationship(Petal.Width, Sepal.Width, color=Species)

```

# `...` is friggin' sweet

```{r}

props = function(data, ...) {
  data %>%
  count(...) %>%
  mutate(prop = n / sum(n), .keep="unused", by=...)
}

iris %>%
  props(Species) %>%
  kable
```

# `...` is friggin' sweet

```{r, fig.width=6}
drop_extreme = function(data, ..., q_lo=0, q_hi=0.95) {
    data %>%
        drop_na(...) %>%
        filter(if_all(c(...), ~
            between(.x, quantile(.x, q_lo), quantile(.x, q_hi))
    ))
}

# <- is for sheeple
bad_iris = iris %>% add_row(Petal.Length=0., Petal.Width=30)

bad_plot = bad_iris %>%
    plot_relationship(Petal.Width, Petal.Length)

good_plot = bad_iris %>%
    drop_extreme(Petal.Width, Petal.Length, q_lo=.025, q_hi=.975) %>%
    plot_relationship(Petal.Width, Petal.Length)

bad_plot + good_plot

```

# `...` is friggin' sweet

- c.f. `function(...args)` in javascript or `def foo(*args, **kwargs)` in python
- the general name for this is "varargs"
- see [Hadley's book](https://adv-r.hadley.nz/functions.html?q=...#fun-dot-dot-dot) for deets


# Another (more composable) approach for reusing plotting code

```{r fig.width=6}
iris_pal = scale_colour_manual(values=c(
    setosa = "#842EBD",
    versicolor = "#CA26E2",
    virginica = "#EB75F5"
))

no_legend = theme(legend.position="none")

point_lm = list(
    geom_point(alpha=0.5),
    stat_smooth(method="lm",  alpha=0.2)
)

iris %>% ggplot(aes(Petal.Width, Petal.Length)) + point_lm +
iris %>% ggplot(aes(Petal.Width, Petal.Length, color=Species)) + point_lm + iris_pal + no_legend
```

# The best of both worlds??

```{r}
bars = function(..., err_width=.1) {
    list(
        stat_summary(fun=mean, geom="bar", ...),
        stat_summary(fun.data=mean_cl_boot, geom="errorbar", width=err_width)
    )
}

iris %>% ggplot(aes(Species, Petal.Width, fill=Species)) +
    bars(width=.7, err_width=0.2, alpha=0.5) +
    iris_pal + no_legend
```

```{r include=FALSE}
robust = function(f, min_n=5) {
    function(x) {
        if (length(x) >= min_n) {
            f(x)
        } else {
            NaN
        }
    }
}

points = function(min_n=5, ...) stat_summary(fun.data=robust(mean_cl_boot, min_n), ...)

points_line = function(min_n=5, ...) {
    list(
        stat_summary(fun.data=robust(mean_cl_boot, min_n), ...),
        stat_summary(fun=robust(mean, min_n), geom="line", ...)
    )
}
```

# More examples...

```{r}
# see code for robust definition -- it drops points with too few observations
bin_points = function(bins, min_n=5, ...) {
    stat_summary_bin(fun.data=robust(mean_cl_boot, min_n), bins=bins, ...)
}

bin_points_line = function(bins, min_n=5, ...) {
    list(
        stat_summary_bin(fun.data=robust(mean_cl_boot, min_n), bins=bins, ...),
        stat_summary_bin(fun=robust(mean, min_n), bins=bins, ..., geom="line")
    )
}

iris %>% ggplot(aes(Petal.Width, Petal.Length)) +
  bin_points_line(bins=7)

```

# More examples...

```{r, fig.width=9}

linear_fit = function(alpha=0.2, ...) stat_smooth(method="lm",  alpha=alpha, ...)
logistic_fit = function(alpha=0.2, ...) stat_smooth(method="glm", method.args = list(family="binomial"), alpha=alpha, ...)
gam_fit = function(k=-1, ...) stat_smooth(method = "gam", formula = y ~ s(x, bs = "cs", k=k), ...)

p = iris %>% ggplot(aes(Petal.Width, Petal.Length))
p + bin_points(7) + linear_fit() +
p + bin_points(7) + gam_fit() +
p + bin_points(7) + gam_fit(k=3, color="dodgerblue")
```


# Bonus: Automatically formatting plot labels

See the source code for `pretty_labels` defintion.

```{r include=FALSE}
element_text_transform <- function(..., transform = identity) {
    elem <- element_text(...)
    elem$transform <- rlang::as_function(transform)
    class(elem) <- c("element_text_transform", class(elem))
    elem
}

element_grob.element_text_transform <- function(element, label = "", ...) {
    label <- element$transform(label)
    NextMethod()
}

# direct mapping from variable name to label
pretty_names = list(
    pid = "Participant"
)

# regex replacements applied to variable names
pretty_replacements = c(
    "1 ?\\* ?" = "",
    "factor\\((.*)\\)" = "\\1",
    "_" = " ",
    "\\." = " ",
    "\\bn\\b" = "#",
    "\\bprop\\b" = "proportion"
)

# additional substitutions performed after capitalization
pretty_fixes = c(
    "Rt" = "RT"
)

pretty_name = function(lab) {
    if (lab %in% names(pretty_names)) return(pretty_names[[lab]])
    if (str_detect(lab, "β") || str_detect(lab, "γ")) return(lab)
    if (nchar(lab) == 1) return(lab)
    lab %>%
        str_replace_all(pretty_replacements) %>%
        # str_to_title %>%   # your choice
        str_to_lower %>%
        str_replace_all(pretty_fixes) %>%
        gsub("(?!^)\\b(A|An|And|As|At|But|By|For|In|Nor|Of|On|Or|Per|The|To|Up|Vs)\\b", "\\L\\1", ., perl=TRUE)
}


pretty_name_compact = function(lab) {
    name = pretty_name(lab)
    if (str_detect(name, " ") && str_length(name) > 10) {
        name = str_replace_all(name, " ", "\n")
    }
    name
}

pretty_labels = theme(
  axis.title.x = element_text_transform(transform = pretty_name),
  axis.title.y = element_text_transform(transform = pretty_name),
  legend.title = element_text_transform(transform = pretty_name_compact),
  strip.text = element_text_transform(transform = pretty_name)
)
```

```{r, fig.width=9}
p = iris %>% ggplot(aes(Petal.Width, Sepal.Width, color=Species)) +
  geom_point(alpha=0.3) + linear_fit() +
  iris_pal

p + (p + pretty_labels)
```

# Bonus: simplifying regression syntax

See the source code for `regress` defintion.

```{r include=FALSE}
library(formula.tools)
library(broom)
library(glue)
library(lme4)
library(broom.mixed)

regress = function(data, form, logistic=F, mixed=F, group_var=NA, intercept=T, standardize=F) {
    preds = paste(get.vars(rhs(form)), collapse=" + ")
    data = tibble(data)  # to not mutate
    if (standardize) {
        for (k in get.vars(form)) {
            if (!is.binary(data[[k]])) {
                data[[k]] = zscore(data[[k]])
            }
        }
    }
    model = if (mixed || !is.na(group_var)) {
        if (!is.na(group_var)) {
            form = as.formula(glue("{form} + {1*intercept} + ({preds} + {1*intercept} || {group_var})"))
        }
        if (logistic) {
            glmer(form, family=binomial, data=data)
        } else {
            lmer(form, data=data)
        }
    } else {
        if (logistic) {
            glm(form, family=binomial, data=data)
        } else {
            lm(form, data=data)
        }
    }
    model
}
```


```{r}
iris %>%
  filter(Species == "versicolor") %>%
  regress(Sepal.Width ~ Petal.Width) %>%
  tidy %>% kable

```

# Bonus: simplifying regression syntax (logistic)

```{r}

iris %>%
  regress(Species == "versicolor" ~ Sepal.Width + Petal.Width, logistic=T) %>%
  tidy %>% kable
```

# Bonus: simplifying regression syntax (mixed effects)

```{r}

iris %>%
  mutate(island = rep(1:15, 10)) %>%
  regress(Species == "versicolor" ~ Sepal.Width + Petal.Width, logistic=T, group_var="island") %>%
  tidy %>% kable
```